\chapter{Acronyms and Definitions}
\label{ch:acronyms-and-definitions}

\section{Acronyms}

\begin{description}[style=nextline]
  \item[AI] Artificial Intelligence
  \item[AHP] Analytic Hierarchy Process
  \item[API] Application Programming Interface
  \item[BI] Business Intelligence
  \item[CDC] Change Data Capture
  \item[CDSS] Clinical Decision Support System
  \item[CNN] Convolutional Neural Network
  \item[CRM] Customer Relationship Management
  \item[DBSCAN] Density-Based Spatial Clustering of Applications with Noise
  \item[DL] Deep Learning
  \item[DNN] Deep Neural Network
  \item[DSS] Decision Support System
  \item[ELT] Extract, Load, Transform
  \item[ERP] Enterprise Resource Planning
  \item[ETL] Extract, Transform, Load
  \item[GDSS] Group Decision Support System
  \item[GPU] Graphics Processing Unit
  \item[IoT] Internet of Things
  \item[KPI] Key Performance Indicator
  \item[LSTM] Long Short-Term Memory
  \item[MCDA] Multi-Criteria Decision Analysis
  \item[MIS] Management Information System
  \item[ML] Machine Learning
  \item[NLP] Natural Language Processing
  \item[OLAP] Online Analytical Processing
  \item[OLTP] Online Transaction Processing
  \item[PCA] Principal Component Analysis
  \item[RNN] Recurrent Neural Network
  \item[SVM] Support Vector Machine
  \item[XAI] Explainable Artificial Intelligence
\end{description}

\section{Core Definitions}

\begin{description}[style=nextline]
  \item[Decision Support System (DSS)] A computer-based system that supports
  semi-structured and unstructured decision making through data, models,
  knowledge, and interactive interfaces.
  \item[Structured decision] A routine decision with a well-defined procedure.
  \item[Semi-structured decision] A decision that combines formal analysis with
  human judgment.
  \item[Unstructured decision] A novel or complex decision with no complete
  predefined procedure.
  \item[Model management] The organization, validation, deployment, monitoring,
  and governance of analytical models used by a DSS.
  \item[Business Intelligence] Technologies and processes used to integrate,
  analyze, and visualize data for decision making.
  \item[Expert system] A knowledge-driven system using explicit rules and an
  inference engine to provide recommendations.
  \item[Human-in-the-loop] A design pattern in which humans review, approve, or
  override algorithmic recommendations.
  \item[Calibration] The degree to which predicted probabilities match observed
  outcome frequencies.
  \item[Data leakage] Use of information during model development that would not
  be available at decision time.
  \item[Cold start] A recommendation problem caused by limited history for new
  users or items.
  \item[Sensitivity analysis] Examination of how outputs change when inputs,
  assumptions, or weights are varied.
  \item[Audit trail] A documented record of data, model/rule versions, and
  actions used for accountability.
  \item[Fairness] The requirement that DSS decisions and recommendations avoid
  unjustified systematic disadvantage across groups.
  \item[Explainability] The degree to which a human can understand why a system
  produced a particular output or recommendation.
\end{description}

\section{How to Use This Chapter}

This chapter is intended as a compact revision aid. Students can use it before
exams, during slide review, and while preparing projects to ensure that core
terminology remains precise and consistent across the course.
