\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[ML in DSS]{Introduction to Machine Learning in DSS}
\subtitle{Chapter 05}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Learning objectives}
  \begin{itemize}
    \item Identify common ML problem types in DSS.
    \item Explain why data preparation and leakage control matter.
    \item Select evaluation metrics aligned with decisions.
    \item Connect prediction outputs to actual policy choices.
  \end{itemize}
\end{frame}

\begin{frame}{Common ML tasks in DSS}
  \begin{itemize}
    \item Classification for risk or category decisions.
    \item Regression for continuous outcome prediction.
    \item Ranking and recommendation for prioritization.
    \item Anomaly detection and forecasting for operational support.
  \end{itemize}
\end{frame}

\begin{frame}{Data preparation and leakage}
  \begin{itemize}
    \item Data quality shapes model quality.
    \item Feature engineering should respect decision-time availability.
    \item Leakage leads to unrealistically strong validation results.
  \end{itemize}
\end{frame}

\begin{frame}{Evaluation for decisions}
  \begin{itemize}
    \item Accuracy alone is often insufficient.
    \item Use precision, recall, calibration, and cost-sensitive evaluation.
    \item Thresholds should reflect capacity and decision consequences.
  \end{itemize}
\end{frame}

\begin{frame}{From prediction to action}
  \begin{itemize}
    \item A probability score is not yet a decision.
    \item DSS must define thresholds, workflows, and exception handling.
    \item Explainability improves actionability and trust.
  \end{itemize}
\end{frame}

\discussionframe{Why is a highly accurate ML model still insufficient if the
decision threshold and intervention workflow are poorly designed?}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Good ML for DSS begins with good problem framing.
    \item The real objective is decision quality, not model accuracy alone.
  \end{itemize}
\end{frame}
\end{document}
