\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Midterm Review]{Midterm Review and Sample Exam Guidance}
\subtitle{Chapter 06}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Midterm coverage}
  \begin{itemize}
    \item Chapters 01--05 form the conceptual and technical foundation.
    \item Focus on definitions, comparisons, workflows, and design trade-offs.
    \item Expect both theory questions and applied DSS design questions.
  \end{itemize}
\end{frame}

\begin{frame}{What students should revise}
  \begin{itemize}
    \item DSS components, types, and AI-enabled extensions.
    \item Decision-making concepts, bias, risk, and uncertainty.
    \item DSS architecture, deployment, security, and governance.
    \item Model management, ML task types, leakage, and evaluation.
  \end{itemize}
\end{frame}

\begin{frame}{Typical question formats}
  \begin{itemize}
    \item Short definitions and comparisons.
    \item Essay questions about design trade-offs.
    \item Scenario-based questions requiring DSS architecture choices.
    \item Metric and threshold interpretation questions.
  \end{itemize}
\end{frame}

\begin{frame}{Answering strategy}
  \begin{itemize}
    \item Start from the decision problem, not from the tool.
    \item Use precise terminology from the course.
    \item Mention governance, bias, and human oversight where relevant.
    \item Support answers with one concrete application example.
  \end{itemize}
\end{frame}

\discussionframe{If you were writing one essay question for the midterm, what
would it test: conceptual understanding, model judgment, or governance? Why?}

\begin{frame}{Revision checklist}
  \begin{itemize}
    \item Can you distinguish MIS, BI, and DSS?
    \item Can you explain bounded rationality and automation bias?
    \item Can you describe the model lifecycle and leakage risk?
    \item Can you connect prediction outputs to decision thresholds?
  \end{itemize}
\end{frame}
\end{document}
