\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Classification Training]{Training of Classification Models}
\subtitle{Chapter 07}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Learning objectives}
  \begin{itemize}
    \item Structure classification problems for DSS.
    \item Use proper train, validation, and test splits.
    \item Handle imbalance, calibration, and threshold setting.
    \item Connect error analysis to governance and monitoring.
  \end{itemize}
\end{frame}

\begin{frame}{Problem definition and splitting}
  \begin{itemize}
    \item Define labels carefully and respect decision time.
    \item Use train, validation, and test sets appropriately.
    \item Time-aware splits matter in operational DSS.
  \end{itemize}
\end{frame}

\begin{frame}{Baselines and pipelines}
  \begin{itemize}
    \item Start with simple baselines before complex models.
    \item Build preprocessing and modeling pipelines together.
    \item Reproducibility matters for academic and operational use.
  \end{itemize}
\end{frame}

\begin{frame}{Imbalanced learning}
  \begin{itemize}
    \item Accuracy can hide poor minority-class performance.
    \item Use precision, recall, F1, PR curves, and cost-aware metrics.
    \item Class weights, resampling, and thresholding are common tools.
  \end{itemize}
\end{frame}

\begin{frame}{Calibration and thresholding}
  \begin{itemize}
    \item Scores should reflect meaningful probabilities.
    \item Thresholds should depend on cost, benefit, and operational capacity.
    \item Decision curves connect model output to policy value.
  \end{itemize}
\end{frame}

\discussionframe{Why can the best AUC model still be the wrong model to deploy
in a real DSS?}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Classification training is a decision design problem, not only a
    modeling problem.
    \item Calibration, thresholding, and monitoring are deployment-critical.
  \end{itemize}
\end{frame}
\end{document}
