\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Supervised Learning]{Supervised Learning for DSS}
\subtitle{Chapter 08}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Learning objectives}
  \begin{itemize}
    \item Compare major supervised learning algorithms.
    \item Explain feature selection and interpretability choices.
    \item Evaluate models beyond accuracy.
    \item Describe how supervised models are operationalized in DSS.
  \end{itemize}
\end{frame}

\begin{frame}{Algorithm families}
  \begin{itemize}
    \item Linear and generalized linear models.
    \item Decision trees and ensembles.
    \item Support Vector Machines.
    \item Neural networks.
  \end{itemize}
\end{frame}

\begin{frame}{Choosing an algorithm}
  \begin{itemize}
    \item Consider data size, interpretability, latency, and governance.
    \item The most accurate model may not be the most deployable model.
    \item DSS contexts often reward robustness and explainability.
  \end{itemize}
\end{frame}

\begin{frame}{Feature selection and interpretability}
  \begin{itemize}
    \item Feature selection can improve simplicity and reduce leakage risk.
    \item Interpretability methods support trust and contestability.
    \item Trade-offs often exist between transparency and raw performance.
  \end{itemize}
\end{frame}

\begin{frame}{Evaluation beyond accuracy}
  \begin{itemize}
    \item Use metrics tied to decision objectives.
    \item Check calibration and communicate uncertainty.
    \item Evaluate policy performance under resource constraints.
  \end{itemize}
\end{frame}

\discussionframe{When would you deliberately choose a less accurate but more
interpretable model in a DSS?}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Supervised learning choices must align with the decision context.
    \item Operational fit and governance are part of model quality.
  \end{itemize}
\end{frame}
\end{document}
