\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Expert Systems]{Automated Decision Systems and Expert Systems}
\subtitle{Chapter 09}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Learning objectives}
  \begin{itemize}
    \item Explain rule-based and expert-system logic.
    \item Describe forward and backward chaining.
    \item Compare expert systems with ML-based DSS.
    \item Explain why hybrid systems are often preferable.
  \end{itemize}
\end{frame}

\begin{frame}{Knowledge representation}
  \begin{itemize}
    \item Facts represent known statements.
    \item Rules encode IF--THEN logic.
    \item Ontologies structure concepts and relationships.
  \end{itemize}
\end{frame}

\begin{frame}{Inference and conflict resolution}
  \begin{itemize}
    \item Forward chaining starts from facts.
    \item Backward chaining starts from goals.
    \item Conflicts are resolved using priority, specificity, or meta-rules.
  \end{itemize}
\end{frame}

\begin{frame}{Hybrid DSS}
  \begin{itemize}
    \item ML handles pattern recognition and scoring.
    \item Rules encode policy, safety, and compliance constraints.
    \item Hybrid systems improve auditability and guardrail enforcement.
  \end{itemize}
\end{frame}

\begin{frame}{Automation and accountability}
  \begin{itemize}
    \item Advisory DSS and human-in-the-loop designs are common.
    \item High-stakes decisions require documentation and contestability.
    \item Automation bias must be actively managed.
  \end{itemize}
\end{frame}

\discussionframe{For which tasks are explicit rules still better than ML:
compliance, diagnosis, triage, or recommendation? Explain.}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Expert systems remain relevant when knowledge must be explicit.
    \item Hybrid designs are often the strongest practical pattern.
  \end{itemize}
\end{frame}
\end{document}
