\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Unsupervised DSS]{Unsupervised Learning and Recommender Systems for DSS}
\subtitle{Chapter 11}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Learning objectives}
  \begin{itemize}
    \item Explain how unlabeled data supports decisions.
    \item Use clustering, dimensionality reduction, and anomaly detection.
    \item Compare content-based, collaborative, and hybrid recommenders.
    \item Identify recommendation risks such as cold start and popularity bias.
  \end{itemize}
\end{frame}

\begin{frame}{Why unsupervised learning matters}
  \begin{itemize}
    \item Many DSS problems start without reliable labels.
    \item Clustering reveals segments for differentiated action.
    \item Dimensionality reduction supports exploration and visualization.
    \item Anomaly detection highlights cases for review.
  \end{itemize}
\end{frame}

\begin{frame}{From patterns to recommendations}
  \begin{itemize}
    \item Content-based recommendation uses attributes.
    \item Collaborative filtering uses behavior similarity.
    \item Hybrid recommenders combine signals and constraints.
  \end{itemize}
\end{frame}

\begin{frame}{Evaluation}
  \begin{itemize}
    \item Unsupervised methods need stability and usefulness checks.
    \item Recommenders use ranking metrics and online response measures.
    \item Fairness and exposure should be monitored explicitly.
  \end{itemize}
\end{frame}

\begin{frame}{Risks}
  \begin{itemize}
    \item Cold start for new users or items.
    \item Popularity bias and feedback loops.
    \item Clusters that are mathematically valid but operationally meaningless.
  \end{itemize}
\end{frame}

\discussionframe{Would you trust an unsupervised segmentation model enough to
allocate resources automatically? Why or why not?}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Unsupervised learning is useful when it leads to interpretable action.
    \item Recommendation systems must be governed, not only optimized.
  \end{itemize}
\end{frame}
\end{document}
