\documentclass[aspectratio=169,11pt]{beamer}
\input{slides/src/common-preamble.sty}

\title[Acronyms and Terms]{Acronyms and Definitions}
\subtitle{Chapter 14}
\author{Dr.\ Haitham El-Ghareeb}
\institute{Faculty of Computers and Information Sciences}
\date{\today}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Why this chapter matters}
  \begin{itemize}
    \item DSS courses use overlapping technical and managerial vocabulary.
    \item Precise terminology improves exam answers and project reports.
    \item Acronyms help students read research and tool documentation faster.
  \end{itemize}
\end{frame}

\begin{frame}{Core acronyms}
  \begin{multicols}{2}
    \begin{itemize}
      \item DSS
      \item MIS
      \item BI
      \item ML
      \item DL
      \item AI
      \item GDSS
      \item MCDA
      \item AHP
      \item XAI
      \item ETL / ELT
      \item OLTP / OLAP
    \end{itemize}
  \end{multicols}
\end{frame}

\begin{frame}{Definitions students should know}
  \begin{itemize}
    \item Structured, semi-structured, and unstructured decision.
    \item Model management, calibration, and data leakage.
    \item Human-in-the-loop, audit trail, fairness, explainability.
    \item Cold start and sensitivity analysis.
  \end{itemize}
\end{frame}

\begin{frame}{How to revise terminology}
  \begin{itemize}
    \item Group terms by chapter and by theme.
    \item Practice using each term in a short applied sentence.
    \item Distinguish related terms rather than memorizing them in isolation.
  \end{itemize}
\end{frame}

\discussionframe{Pick three acronyms from the course and explain how they are
connected in one coherent DSS example.}

\begin{frame}{Key takeaways}
  \begin{itemize}
    \item Terminology is part of professional competence in DSS.
    \item Definitions become useful only when connected to decisions and cases.
  \end{itemize}
\end{frame}
\end{document}
