\documentclass[11pt,oneside]{book}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{bookmark}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

% Simple placeholders helpers
\newcommand{\placeholder}[1]{%
  \par\noindent\textcolor{gray}{\textit{[Placeholder: #1]}}\par
}

\title{Decision Support Systems: A Modern Approach}
\author{Dr.\ Haitham El-Ghareeb\\
Associate Professor of Information Systems,\\
Faculty of Computers and Information Sciences,\\
Mansoura University, Egypt}
\date{\today}

\begin{document}

\frontmatter
\maketitle
\tableofcontents

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
This book was prepared as a teaching text for a Decision Support Systems (DSS)
course with a practical orientation toward analytics, machine learning, and
organizational decision making. The aim is to bridge classical DSS foundations
with modern AI-enabled practice while keeping the material suitable for course
delivery, assignments, revision, and slide-based lecturing.

The sequence starts from core decision-making concepts, moves through DSS
architecture and model management, and then develops the analytics side of the
course through machine learning, business intelligence, expert systems,
unsupervised methods, and multi-criteria decision support. The final chapters
support revision through essay questions and a compact chapter of acronyms and
definitions.

\chapter*{How to Use This Book}
\addcontentsline{toc}{chapter}{How to Use This Book}
The chapters are designed to work at three levels:
\begin{itemize}
  \item \textbf{Conceptual study}: read the overview, main sections, and summary.
  \item \textbf{Assessment preparation}: use the key terms, MCQs, essay chapter,
  and midterm guidance.
  \item \textbf{Project development}: adapt the suggested DSS projects at the end
  of most chapters.
\end{itemize}

For a standard course sequence, Chapters 1--5 establish foundations, Chapters
7--12 deepen analytical and design capabilities, and Chapters 13--14 support
revision and consolidation.

\mainmatter

% ==========================================================
% Chapters (matching your course-topic list)
% ==========================================================

\include{chapters/ch01-introduction}

\include{chapters/ch02-decision-making-concepts}

\include{chapters/ch03-technologies-environment}

\include{chapters/ch04-model-management}

\include{chapters/ch05-intro-ml-in-dss}

\include{chapters/ch06-midterm}

\include{chapters/ch07-training-classification}

\include{chapters/ch08-supervised-learning}

\include{chapters/ch09-automated-expert-systems}

\include{chapters/ch10-business-intelligence}

\include{chapters/ch11-unsupervised-recommender-systems}

\include{chapters/ch12-mcda}

\include{chapters/ch13-essay-questions}

\include{chapters/ch14-acronyms-definitions}

\backmatter
\chapter*{Appendix A: Python Setup and Tooling}
\addcontentsline{toc}{chapter}{Appendix A: Python Setup and Tooling}
\placeholder{Environment setup, packages, and reproducibility.}

\chapter*{Appendix B: Mathematical Background}
\addcontentsline{toc}{chapter}{Appendix B: Mathematical Background}
\placeholder{Probability, linear algebra, optimization basics.}

\chapter*{Bibliography}
\addcontentsline{toc}{chapter}{Bibliography}
\placeholder{Add Bib\TeX{} file later.}

\chapter*{Index}
\addcontentsline{toc}{chapter}{Index}
\placeholder{Index placeholder.}

\end{document}
